---
title: "Package 'odr'"
author: "Zuchao Shen, Ben Kelcey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package 'odr'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
The costs of sampling each additional unit in multilevel experimental studies
vary across levels of hierarchy and treatment conditions due to the 
hierarchical sampling and the delivery of treatment. 
This package is a tool to optimize the 
designs of multilevel experimental studies such that the variances of 
treatment effects are minimized under fixed budget and cost structure, or
the budget is minimized to achieve same level design precision or statistical power.
The optimal sample allocation or optimal design parameters include 

- the optimal sample sizes at each of the levels except the top level because the top-level
sample size will be determined by total budget once all other parameters are decided
- proportion of units to be assigned to treatment condition 
at the level of randomization

This package includes three categorical of functions and they are

- *od* function: This function calculates optimal sample allocation parameters with and without 
constraint(s). For each type of multilevel experimental studies, there is an additional number (and/or
additional letter) to be added to the general function name. For example, the function for 
two-level cluster randomized trials (CRTs) is *od.2*,
the function for 2-level multisite cluster randomized trials is *od.2m*.

- *power* function: This function performs power analyses with and without accommodating costs.
For power analysis accommodating costs, the function calculates required budget and required sample size, 
statistical power, minimum detectable effect size. For conventional power analysis without accommodating costs,
this function calculates required sample size, statistical power, minimum detectable effect size. 
For each type of multilevel experimental studies, there is an additional number (and/or
additional letter) to be added to the general function name. For example, the function for 
two-level cluster randomized trials (CRTs) is *power.2*,
the function for 2-level multisite cluster randomized trials is *power.2m*.

- *re* function: This function calculates relative efficiency values between two designs.


## 1. Function *od* 
Given cost structure (i.e., the costs of sampling each unit at 
different levels and treatment conditions), this function solves 
the optimal sample allocation with and without constraints. 

To solve the optimal sample allocation of a two-level CRT, 
we need the following information

- icc: intraclass correlation coefficient in the population or in each treatment condition
- r12: the proportion of level-1 outcome variance explained by covariates in the population or in each treatment condition
- r22: the proportion of level-2 outcome variance explained by covariates in the population or in each treatment condition
- c1: the cost of sampling each additional level-1 unit in control condition 
- c2: the cost of sampling each additional level-2 unit in control condition 
- c1t: the cost of sampling each additional level-1 unit in treatment condition 
- c2t: the cost of sampling each additional level-2 unit in treatment condition
- m: total fixed budget used to plot the variance curves, default is the cost of sampling 60 level-2 units
across treatment conditions.

### 1.1 Examples
```{r}
library(odr)
```

```{r fig.width = 7, fig.height = 3.5}
 # unconstrained optimal design
myod1 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
              varlim = c(0.01, 0.02))
 # The function by default prints messages of output and plots the variance curves; one can turn off message and specify one or no plot; default m value is the total costs of sampling 60 level-2 units across treatment conditions, m can be explicitly specified.
 # myod1$out # output; 
 # myod1$par # parameters used in the calculation.
```

```{r fig.width = 5, fig.height = 5}
 # constrained optimal design with n = 20
myod2 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
              plot.by = list(p = "p"), n = 20, varlim = c(0.005, 0.030))
 # myod2$out # output
 # myod2$par # parameters used in the calculation.
```

```{r fig.width = 7, fig.height = 3.5}
 # constrained optimal design with p = 0.5
myod3 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
             p = 0.5, varlim = c(0.005, 0.020))
 # myod3$out # output; 
 # myod3$par # parameters used in the calculation.
```

```{r fig.width = 7, fig.height = 3.5}
 # constrained n and p, no calculation performed
myod4 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
              plots = FALSE, n = 20, p = 0.5, varlim = c(0.005, 0.025))
```

### 1.2 Examples for other types of trials
Please see examples in corresponding functions by uncomment below lines.
```{r}
# ?od.1
# ?od.3
# ?od.4
# ?od.2m
# ?od.3m
# ?od.4m
```

## 2. Function *power*
This function by default can perform power analyses accommodating cost structures 
(i.e., cost.model = TRUE), one of 'power', 'm', and 'd' must be NULL. For example,
if 'power' is NULL, the function calculates statistical power 
under a fixed budget and cost structure; if 'd' is NULL, the function
calculates minimum detectable effect size (i.e., d) under a fixed budget and 
desired power level; if 'm' is NULL, the function calculate
required budget (and required sample size) to achieve desired power level to detect a treatment effect.

This function also can conduct conventional power analysis or power analysis wihout 
accommodating cost structures by specifying cost.model = FALSE, the conventional
power analyses include statistical power calculation, MDES calculation, 
and required sample size calculation.

### 2.1 Examples of power analyses accommodating cost structures (cost.model = TRUE)
###### Required budget for desired power
- Required budget calculation
```{r}
mym <- power.2(expr = myod1, d = 0.3, q = 1, power = 0.8)
# mym$out  # m =1702, J = 59
```
- Effects on required budget to maintain same level power when designs depart from the optimal one 
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c(1, 2))
budget <- NULL
nrange <- c(2:50)
for (n in nrange)
  budget <- c(budget, power.2(expr = myod1, constraint = list (n = n), d = 0.3, q = 1, power = 0.8)$out$m)
plot(nrange, budget, type = "l", lty = 1, xlim = c(0, 50), ylim = c(1500, 3500),
     xlab = "Level-1 sample size: n", ylab = "Budget", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
budget <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  budget <- c(budget, power.2(expr = myod1, constraint = list (p = p), d = 0.3, q = 1, power = 0.8)$out$m)
plot(prange, budget, type = "l", lty = 1, xlim = c(0, 1), ylim = c(1500, 7000),
     xlab = "Porportion groups in treatment: p", ylab = "Budget", main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
par(figure)
```

###### Statistical power under a fixed budget
- Power calculation
```{r}
mypower <- power.2(expr = myod1, q = 1, d = 0.3, m = 1702)
# mypower$out  # power = 0.80
```
- Effects on power under same budget when designs depart from the optimal one
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
pwr <- NULL
nrange <- c(2:50)
for (n in nrange)
  pwr <- c(pwr, power.2(expr = myod1, constraint = list (n = n), d = 0.3, q = 1, m = 1702)$out)
plot(nrange, pwr, type = "l", lty = 1, xlim = c(0, 50), ylim = c(0.4, 0.9),
     xlab = "Level-1 sample size: n", ylab = "Power", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
pwr <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  pwr <- c(pwr, power.2(expr = myod1, constraint = list (p = p), d = 0.3, q = 1, m = 1702)$out)
plot(prange, pwr, type = "l", lty = 1, xlim = c(0, 1), ylim = c(0.1, 0.9),
     xlab = "Porportion groups in treatment: p", ylab = "Power",  main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
 par(figure)
```

###### MDES (minimum detectable effect size) under a fixed budget
- MDES calculation
```{r}
mymdes <- power.2(expr = myod1, q = 1, power = 0.80, m = 1702)
# mymdes$out  # d = 0.30
```
- Effects on MDES under same budget when designs depart from the optimal one 
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
MDES <- NULL
nrange <- c(2:50)
for (n in nrange)
  MDES <- c(MDES, power.2(expr = myod1, constraint = list (n = n), power = 0.8, q = 1, m = 1702)$out)
plot(nrange, MDES, type = "l", lty = 1, xlim = c(0, 50), ylim = c(0.3, 0.8),
     xlab = "Level-1 sample size: n", ylab = "MDES", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
MDES <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  MDES <- c(MDES, power.2(expr = myod1, constraint = list (p = p), power = 0.8, q = 1, m = 1702)$out)
plot(prange, MDES, type = "l", lty = 1, xlim = c(0, 1), ylim = c(0.3, 0.8),
     xlab = "Porportion groups in treatment: p", ylab = "MDES", main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
 par(figure)
```

### 2.2 Examples of conventional power analyses (cost.model = FALSE)
```{r}
# Required level-2 sample size calculation
myJ <- power.2(cost.model = FALSE, expr = myod1, d = 0.3, q = 1, power = 0.8)
# above experssion takes parameters and outputs from od.2 function. Equivalently, each parameter can be explicitly specified.
# myJ <- power.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
#                     cost.model = FALSE, n = 9, p = 0.33, d = 0.3, q = 1, power = 0.8)
myJ$out  # J = 59

# Power calculation
mypower1 <- power.2(cost.model = FALSE, expr = myod1, J = 59, d = 0.3, q = 1)
mypower1$out  # power = 0.80

# MDES calculation
mymdes1 <- power.2(cost.model = FALSE, expr = myod1, J = 59, power = 0.8, q = 1)
mymdes1$out  # d = 0.30
```
### 2.3 Examples of conventional power curves
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
pwr <- NULL
mrange <- c(300:3000)
for (m in mrange)
  pwr <- c(pwr, power.2(expr = myod1, d = 0.3, q = 1, m = m)$out)
plot(mrange, pwr, type = "l", lty = 1, xlim = c(300, 3000), ylim = c(0, 1),
     xlab = "Budget", ylab = "Power", main = "", col = "black")
 abline(v = 1702, lty = 2, col = "Blue")
 
pwr <- NULL
Jrange <- c(4:100)
for (J in Jrange)
  pwr <- c(pwr, power.2(expr = myod1, cost.model = FALSE, d = 0.3, q = 1, J = J)$out)
plot(Jrange, pwr, type = "l", lty = 1, xlim = c(4, 100), ylim = c(0, 1),
     xlab = "Level-2 sample size: J", ylab = "Power", main = "", col = "black")
 abline(v = 59, lty = 2, col = "Blue")
par(figure)
```

### 2.4 Examples for other types of trials
Please see examples in corresponding functions by uncomment below lines (e.g., type ?power.3).
```{r}
# ?power.1
# ?power.3
# ?power.4
# ?power.2m
# ?power.3m
# ?power.4m
```

## 3. Function *re* 
Calculate the relative efficiency (RE) of two designs, this function uses the returns from *od* function

### 3.1 Examples
Based on above examples in *od* function, calculate the relative efficiency
```{r}
# relative efficiency (RE) of constrained design comparing with the unconstrained design
myre <- re(od = myod1, subod= myod2)
myre$re # get the output (i.e., RE = 0.88)

# relative efficiency (RE) constrained design comparing with the unconstrained optimal one
myre <- re(od = myod1, subod= myod3)

# relative efficiency (RE) constrained design comparing with the unconstrained optimal one
myre <- re(od = myod1, subod= myod4)


```
### 3.2 Examples for other types of trials
Please see examples in corresponding functions by uncomment the below lines.
```{r}
# ?od.1
# ?od.2
# ?od.3
# ?od.4
# ?od.2m
# ?od.3m
# ?od.4m
```
