---
title: "Package 'odr'"
author: "Zuchao Shen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package 'odr'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
The costs of sampling each additional unit in multilevel experimental studies
vary across levels of hierarchy and treatment conditions due to the 
hierarchical sampling and the delivery of treatment. 
This package is a tool to optimize the 
designs of multilevel experimental studies such that the variances of 
treatment effects are minimized under fixed budget and cost structure, or
the budget is mimimized to achieve same level design precision. The 
optimal sample allocation or optimal design parameters include 

- the optimal sample sizes at each of the levels except the top level because the top
level sample size will be determined by total budget once all other parameters are decided
- proportion of units assigned to treatment condition 
at the level of randomization

This package includes two categorical of functions and they are

- od
- power

For each type of multilevel experimental studies, there is additional number(s) and/or
letter(s) to be added to the general function name. For example,
the two functions for 2-level group randomized trials (GRTs) with
treatment randomization at the level-2 are od.2 and power.2; the two 
functions for 2-level multisite randomized trials with
treatment randomization at level-1 are od.2m and power.2m

## Function: od 
Given cost structure (i.e., the costs of sampling each unit at 
different levels and treatment conditions), this function solves 
the optimal sample allocation with and without constraints. 

To solve the optimal sample allocation of 2-level GRTs, 
we need the following information

- ICC: intraclass correlation coefficient in the population or in each treatment condtion
- R12: the level-1 outcome variance explained by covariates in the population or in each treatment condtion
- R22: the level-2 outcome variance explained by covariates in the population or in each treatment condtion
- c1: the cost of sampling each additional level-1 unit in control condition 
- c2: the cost of sampling each additional level-2 unit in control condition 
- c1t: the cost of sampling each additional level-1 unit in treatment condition 
- c2t: the cost of sampling each additional level-2 unit in treatment condition
- m: total fixed budget used to plot the variance curves, default is the cost of sampling 60 level-2 units
across treatment conditions.

### Examples
```{r}
library(odr)
```

```{r fig.width = 7, fig.height = 3.5}
 # unconstrained optimal design
myod1 <- od.2(ICC = 0.2, R12 = 0.5, R22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
              varlim = c(0.01, 0.02))
 # the function by default prints messages of output and plots the variance curves; one can turn off message and specify one or no plot; default m value is the total costs of sampling 60 level-2 units across treatment conditions, m can be explicitly specified.
 # myod1$out for output; myod1$par for parameters used in the calculation.
```

```{r fig.width = 5, fig.height = 5}
 # constrained optimal design with n = 20
myod2 <- od.2(ICC = 0.2, R12 = 0.5, R22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
              plot.by = list(p = "p"), n = 20, varlim = c(0.005, 0.030))
 # myod2$out for output; myod2$par for parameters used in the calculation.
```

```{r fig.width = 7, fig.height = 3.5}
 # constrained optimal design with p = 0.5
myod3 <- od.2(ICC = 0.2, R12 = 0.5, R22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
             p = 0.5, varlim = c(0.005, 0.020))
 # myod3$out for output; myod3$par for parameters used in the calculation.
```

```{r fig.width = 7, fig.height = 3.5}
 # constrained n and p, no calculation performed
myod4 <- od.2(ICC = 0.2, R12 = 0.5, R22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
              plot = FALSE, n = 20, p = 0.5, varlim = c(0.005, 0.025))
```

## Function: power
This function performs the calculation directly from 
a cost-analysis perspective. For example, this function can provide statistical power
calculation under fixed budget and cost structure, MDES calculation under fixed buget
and cost structure with desired power level, budget calculation with desired 
power to detect the treatment effect. This function also provides conventional 
statistical power analyses
of multilevel experimental studies (e.g., power calculation, 
minimum detectable effect size (MDES) calculation, required
sample size calculation). 

This function provides power analyses with cost model on With default 
(i.e., cost.model = T), one of 'power', 'm', and 'es' must be NULL. For example,
if 'power' is NULL, the function calculates statistical power of treatment effect
under fixed budget and cost structure; if 'es' is NULL, the fuction
calculate minimum detectable effect size (i.e., es) under fixed budget and 
desired power level; if 'm' is NULL, the function calculate
required budget to achieve desired power level to detect a treatment effect.
When cost model is off (i.e., cost.model = F), this function provides conventional
power analyses that can calculate statistical power, MDES, or required sample size.

### Examples of power analyses accommodating costs and budget (cost.model = T)
###### Required budget for desired power
- Required budget calculation
```{r}
mym <- power.2(expr = myod1, es = 0.3, q = 1, power = 0.8)
# mym$out  # m =1702
```
- Effects on required budget when designs depart from the optimal one 
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c(1, 2))
budget <- NULL
nrange <- c(2:50)
for (n in nrange)
  budget <- c(budget, power.2(expr = myod1, constraint = list (n = n), es = 0.3, q = 1, power = 0.8)$out)
plot(nrange, budget, type = "l", lty = 1, xlim = c(0, 50), ylim = c(1500, 3500),
     xlab = "Level-1 sample size: n", ylab = "Budget", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
budget <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  budget <- c(budget, power.2(expr = myod1, constraint = list (p = p), es = 0.3, q = 1, power = 0.8)$out)
plot(prange, budget, type = "l", lty = 1, xlim = c(0, 1), ylim = c(1500, 7000),
     xlab = "Porportion groups in treatment: p", ylab = "Budget", main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
par(figure)
```

###### Statistical power under fixed budget
- Power calculation
```{r}
mypower <- power.2(expr = myod1, q = 1, es = 0.3, m = 1702)
# mypower$out  # power = 0.80
```
- Effects on power when designs depart from the optimal one
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
pwr <- NULL
nrange <- c(2:50)
for (n in nrange)
  pwr <- c(pwr, power.2(expr = myod1, constraint = list (n = n), es = 0.3, q = 1, m = 1702)$out)
plot(nrange, pwr, type = "l", lty = 1, xlim = c(0, 50), ylim = c(0.4, 0.9),
     xlab = "Level-1 sample size: n", ylab = "Power", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
pwr <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  pwr <- c(pwr, power.2(expr = myod1, constraint = list (p = p), es = 0.3, q = 1, m = 1702)$out)
plot(prange, pwr, type = "l", lty = 1, xlim = c(0, 1), ylim = c(0.1, 0.9),
     xlab = "Porportion groups in treatment: p", ylab = "Power",  main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
 par(figure)
```

###### MDES under fixed budget
- MDES calculation
```{r}
mymdes <- power.2(expr = myod1, q = 1, power = 0.80, m = 1702)
# mymdes$out  # MDES = 0.30
```
- Effects on MDES when designs depart from the optimal one 
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
MDES <- NULL
nrange <- c(2:50)
for (n in nrange)
  MDES <- c(MDES, power.2(expr = myod1, constraint = list (n = n), power = 0.8, q = 1, m = 1702)$out)
plot(nrange, MDES, type = "l", lty = 1, xlim = c(0, 50), ylim = c(0.3, 0.8),
     xlab = "Level-1 sample size: n", ylab = "MDES", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
MDES <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  MDES <- c(MDES, power.2(expr = myod1, constraint = list (p = p), power = 0.8, q = 1, m = 1702)$out)
plot(prange, MDES, type = "l", lty = 1, xlim = c(0, 1), ylim = c(0.3, 0.8),
     xlab = "Porportion groups in treatment: p", ylab = "MDES", main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
 par(figure)
```

### Examples of conventional power analyses (cost.model = FALSE)
```{r}
# Required level-2 sample size calculation
myJ <- power.2(cost.model = FALSE, expr = myod1, es = 0.3, q = 1, power = 0.8)
# above experssion takes parameters and outputs from od.2 function. Equivalently, each parameter can be explicitly specified.
# myJ <- power.2(ICC = 0.2, R12 = 0.5, R22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
#                     cost.model = FALSE, n = 9, p = 0.33, es = 0.3, q = 1, power = 0.8)
myJ$out  # J = 59

# Power calculation
mypower1 <- power.2(cost.model = FALSE, expr = myod1, J = 59, es = 0.3, q = 1)
mypower1$out  # power = 0.80

# MDES calculation
mymdes1 <- power.2(cost.model = FALSE, expr = myod1, J = 59, power = 0.8, q = 1)
mymdes1$out  # es = 0.30
```
### Examples of conventional power plots
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
pwr <- NULL
mrange <- c(300:3000)
for (m in mrange)
  pwr <- c(pwr, power.2(expr = myod1, es = 0.3, q = 1, m = m)$out)
plot(mrange, pwr, type = "l", lty = 1, xlim = c(300, 3000), ylim = c(0, 1),
     xlab = "Budget", ylab = "Power", main = "", col = "black")
 abline(v = 1702, lty = 2, col = "Blue")
 
pwr <- NULL
Jrange <- c(4:100)
for (J in Jrange)
  pwr <- c(pwr, power.2(expr = myod1, cost.model = FALSE, es = 0.3, q = 1, J = J)$out)
plot(Jrange, pwr, type = "l", lty = 1, xlim = c(4, 100), ylim = c(0, 1),
     xlab = "Level-2 sample size: J", ylab = "Power", main = "", col = "black")
 abline(v = 59, lty = 2, col = "Blue")
par(figure)
```


